title: "Draft: Results"

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(broom)
library(scales)

theme_set(theme_minimal())
options(scipen = 999)

degF_to_degC <- function(f) (f - 32) * 5 / 9
threshold_85F_C <- degF_to_degC(85)

# 1. Load Gulf water temperature data (Dauphin Island, DPHA1)
water_raw <- read_table(
  "water_temps.txt",
  col_names = c(
    "year", "month", "day", "hour", "minute",
    "wdir", "wspd", "gst", "wvht", "dpd", "apd", "mwd",
    "pres", "atmp", "wtmp", "dewp", "vis", "tide"
  ),
  skip = 2
)

water <- water_raw %>%
  mutate(
    datetime_local = make_datetime(
      year = year,
      month = month,
      day = day,
      hour = hour,
      min = minute,
      tz = "America/Chicago"
    ),
    # Sentinel code for missing water temps is 99.00
    wtmp = na_if(wtmp, 99.00),
    wtmp_F = wtmp * 9 / 5 + 32,
    exceed_85F = wtmp_F > 85
  ) %>%
  filter(!is.na(wtmp_F))

# Aggregate to hourly resolution for alignment with demand data
water_hourly <- water %>%
  mutate(timestamp_hour = floor_date(datetime_local, "hour")) %>%
  group_by(timestamp_hour) %>%
  summarise(
    wtmp_F = mean(wtmp_F, na.rm = TRUE),
    exceed_85F = any(exceed_85F, na.rm = TRUE),
    .groups = "drop"
  )

# 2. Load hourly electricity demand data
# Assumption: the CSV has already been filtered to the balancing authority
# that includes the Dauphin Island plant.
elec_raw <- read_csv("electrical_usage.csv") %>%
  clean_names()

elec <- elec_raw %>%
  mutate(
    timestamp_hour = ymd_hms(local_time_at_end_of_hour, tz = "America/Chicago")
  ) %>%
  arrange(timestamp_hour)

# 3. Merge water temperature and demand
data_merged <- elec %>%
  left_join(water_hourly, by = "timestamp_hour") %>%
  mutate(
    month = month(timestamp_hour, label = TRUE, abbr = TRUE),
    doy = yday(timestamp_hour)
  )

# Define peak-demand indicator as top 10% of demand hours
peak_threshold <- quantile(data_merged$demand_mw, probs = 0.90, na.rm = TRUE)

data_merged <- data_merged %>%
  mutate(
    peak_demand = demand_mw >= peak_threshold
  )

# 4. Core descriptive summaries
water_summary <- water %>%
  summarise(
    n_obs = n(),
    start_date = min(datetime_local, na.rm = TRUE),
    end_date = max(datetime_local, na.rm = TRUE),
    mean_F = mean(wtmp_F, na.rm = TRUE),
    sd_F = sd(wtmp_F, na.rm = TRUE),
    min_F = min(wtmp_F, na.rm = TRUE),
    p50_F = median(wtmp_F, na.rm = TRUE),
    p95_F = quantile(wtmp_F, 0.95, na.rm = TRUE),
    max_F = max(wtmp_F, na.rm = TRUE),
    prop_exceed_85F = mean(exceed_85F, na.rm = TRUE)
  )

demand_summary <- data_merged %>%
  summarise(
    n_hours = sum(!is.na(demand_mw)),
    start_hour = min(timestamp_hour, na.rm = TRUE),
    end_hour = max(timestamp_hour, na.rm = TRUE),
    mean_MW = mean(demand_mw, na.rm = TRUE),
    sd_MW = sd(demand_mw, na.rm = TRUE),
    min_MW = min(demand_mw, na.rm = TRUE),
    p50_MW = median(demand_mw, na.rm = TRUE),
    p95_MW = quantile(demand_mw, 0.95, na.rm = TRUE),
    max_MW = max(demand_mw, na.rm = TRUE),
    prop_peak_hours = mean(peak_demand, na.rm = TRUE)
  )

exceed_metrics <- data_merged %>%
  summarise(
    n_hours_with_temp = sum(!is.na(exceed_85F)),
    n_exceed_hours = sum(exceed_85F, na.rm = TRUE),
    prop_exceed = mean(exceed_85F, na.rm = TRUE),
    n_peak_exceed_hours = sum(exceed_85F & peak_demand, na.rm = TRUE),
    prop_exceed_in_peak = mean(exceed_85F[peak_demand], na.rm = TRUE),
    prop_exceed_in_offpeak = mean(exceed_85F[!peak_demand], na.rm = TRUE)
  )

# Translate observed probabilities into expected annual and 5‑year hours,
# assuming that 2025 conditions are representative of the next 5 years.
hours_per_year <- 365 * 24
annual_exceed_hours <- exceed_metrics$n_exceed_hours / exceed_metrics$n_hours_with_temp * hours_per_year
annual_exceed_days <- annual_exceed_hours / 24
five_year_exceed_hours <- 5 * annual_exceed_hours
five_year_exceed_days <- 5 * annual_exceed_days

annual_peak_exceed_hours <- exceed_metrics$n_peak_exceed_hours / exceed_metrics$n_hours_with_temp * hours_per_year
annual_peak_exceed_days <- annual_peak_exceed_hours / 24

# 5. Simple models of exceedance risk
model_crude <- glm(exceed_85F ~ 1, family = binomial, data = data_merged)
model_peak <- glm(exceed_85F ~ peak_demand, family = binomial, data = data_merged)
model_adjusted <- glm(exceed_85F ~ peak_demand + month, family = binomial, data = data_merged)

model_table <- list(
  `Crude (overall probability)` = model_crude,
  `Peak vs non‑peak` = model_peak,
  `Peak + month (adjusted)` = model_adjusted
) %>%
  map_dfr(
    ~ tidy(.x, exponentiate = TRUE, conf.int = TRUE),
    .id = "model"
  ) %>%
  filter(term != "(Intercept)" | model == "Crude (overall probability)") %>%
  mutate(
    estimate = round(estimate, 2),
    conf.low = round(conf.low, 2),
    conf.high = round(conf.high, 2)
  )
```

## 1. Sample characteristics and descriptive statistics

We analyzed Gulf water temperatures measured at the Dauphin Island, AL buoy and hourly electricity demand from the regional grid operator.
The water-temperature record includes `r comma(water_summary$n_obs)` half-hour observations from
`r format(water_summary$start_date, "%B %d, %Y")` through
`r format(water_summary$end_date, "%B %d, %Y")`.
The demand dataset contributes `r comma(demand_summary$n_hours)` hourly observations from
`r format(demand_summary$start_hour, "%B %d, %Y")` through
`r format(demand_summary$end_hour, "%B %d, %Y")`.

Data cleaning steps are summarized below:

- **Water temperature**: limited to valid `WTMP` values, converted from °C to °F, and aggregated from 30‑minute to hourly averages to align with demand data. Sentinel missingness codes (e.g., 99.00) were treated as missing.
- **Electricity demand**: limited to observed hourly demand (`demand_mw`), with a peak-demand indicator defined as the top 10% of demand hours.
- **Merged analysis file**: constructed by joining hourly Gulf temperatures to hourly demand on local timestamp.

### 1.1 Descriptive statistics: water temperature

On average, Gulf water temperature during the 2025 observation window was
**`r round(water_summary$mean_F, 1)`°F** (SD `r round(water_summary$sd_F, 1)`°F),
with a median of `r round(water_summary$p50_F, 1)`°F and a maximum of
`r round(water_summary$max_F, 1)`°F.
High-temperature conditions relevant for plant operations were not rare:
**`r percent(water_summary$prop_exceed_85F)`** of half‑hour readings exceeded the
85°F design assumption.

```{r}
water_by_month <- water %>%
  mutate(month = month(datetime_local, label = TRUE, abbr = TRUE)) %>%
  group_by(month) %>%
  summarise(
    n = n(),
    mean_F = mean(wtmp_F, na.rm = TRUE),
    p95_F = quantile(wtmp_F, 0.95, na.rm = TRUE),
    max_F = max(wtmp_F, na.rm = TRUE),
    prop_exceed_85F = mean(exceed_85F, na.rm = TRUE)
  )

knitr::kable(
  water_by_month,
  caption = "Table 1. Monthly Gulf water temperatures and share of time above 85°F (Dauphin Island, 2025).",
  digits = 2
)
```

::: callout-note

Numbering is the same as in draft-data-doc.qmd -- don't do that job for quarto.

Your code doesn't work -- make_datetime() doesn't like the character inputs you're giving it. You probably want to transform things before calling that function.

In general, you need a lot more words and to connect your figures and tables into the text using labels/references as demonstrated in class.

You need to explain the assumptions of your models and what they tell us, specifically. Does it matter that you're controlling for seasonality when that's an inherent part of temperature and demand fluctuation? What do your crude and adjusted models actually mean in reality? You need to spend a lot more time on what the meaning is here, and not just on the fact that you fit the model and ran a test or calculated an odds ratio. Remember, your audience is a manager who hasn't had this material and doesn't care what an odds ratio is. Focus on practical interpretation.

I'm not fixing your code to get everything working, so I haven't examined your graphs yet, but please make sure you're using good color scales, etc. -- I'm happy to take a look once you fix the code.

:::

Monthly summaries show that exceedances of 85°F are concentrated in the late-summer and early‑fall months, when both the 95th percentile and maximum temperatures peak.
This seasonality is important for planning maintenance and contingency capacity.

### 1.2 Descriptive statistics: electricity demand and peak periods

Hourly demand over the same period averaged **`r round(demand_summary$mean_MW, 0)` MW**
(SD `r round(demand_summary$sd_MW, 0)` MW), with a typical hour (median) at
`r round(demand_summary$p50_MW, 0)` MW and a maximum of
`r round(demand_summary$max_MW, 0)` MW.
By construction, approximately **`r percent(demand_summary$prop_peak_hours)`**
of hours are classified as **peak‑demand** hours.

```{r}
demand_by_month <- data_merged %>%
  group_by(month) %>%
  summarise(
    n_hours = sum(!is.na(demand_mw)),
    mean_MW = mean(demand_mw, na.rm = TRUE),
    p95_MW = quantile(demand_mw, 0.95, na.rm = TRUE),
    max_MW = max(demand_mw, na.rm = TRUE)
  )

knitr::kable(
  demand_by_month,
  caption = "Table 2. Monthly electricity demand (MW) in the plant’s region (2025).",
  digits = 0
)
```

Demand exhibits the expected intra‑annual pattern, with higher loads during hot months and lower loads during shoulder seasons.
Because both water temperature and demand peak in warm months, there is potential for coincident operational stress.

### 1.3 Time-series overview

```{r}
ggplot(water_hourly, aes(x = timestamp_hour, y = wtmp_F)) +
  geom_line(color = "#2c7bb6") +
  geom_hline(yintercept = 85, linetype = "dashed", color = "red") +
  labs(
    x = "Date",
    y = "Hourly Gulf water temperature (°F)",
    title = "Figure 1. Hourly Gulf water temperature at Dauphin Island (2025)",
    subtitle = "Red line denotes the plant’s 85°F design assumption."
  )
```

The time series confirms a pronounced warm‑season hump, with extended stretches where hourly average water temperature exceeds 85°F.
These stretches translate directly into periods when the plant would be forced to operate at reduced capacity under current design constraints.

## 2. Primary analysis: probability and frequency of exceeding 85°F

### 2.1 Overall probability of exceedance

At the hourly level (after aggregating 30‑minute readings), the plant would have faced
reduced‑capacity conditions in **`r percent(exceed_metrics$prop_exceed)`** of observed hours.
Scaling this observed proportion to a full year implies that, under 2025‑like conditions, the plant should expect approximately
**`r round(annual_exceed_hours, 0)` hours** (about
**`r round(annual_exceed_days, 1)` days**) per year above 85°F.
Over a 5‑year planning horizon, this corresponds to roughly
**`r round(five_year_exceed_hours, 0)` hours** (about
**`r round(five_year_exceed_days, 1)` days) of required reduced‑capacity operation,
assuming no further warming trend beyond 2025.

```{r}
exceed_by_month <- data_merged %>%
  group_by(month) %>%
  summarise(
    n_hours = sum(!is.na(exceed_85F)),
    n_exceed = sum(exceed_85F, na.rm = TRUE),
    prop_exceed = n_exceed / n_hours
  )

knitr::kable(
  exceed_by_month,
  caption = "Table 3. Monthly share of hours with Gulf water temperature above 85°F (hourly averages).",
  digits = 3
)
```

Monthly probabilities highlight that exceedances are highly seasonal, with very low risk in winter and spring and substantial risk concentrated in the hottest months.
Operationally, this means that reduced‑capacity events are not evenly scattered throughout the year but clustered into specific high‑risk seasons.

### 2.2 Crude and adjusted models of exceedance risk

We fit simple logistic regression models with the binary outcome
“hourly water temperature above 85°F.”

- **Crude model**: intercept‑only, estimating the overall probability of exceedance.
- **Peak‑only model**: includes an indicator for peak‑demand hours.
- **Adjusted model**: includes peak‑demand status and month indicators to account for seasonal confounding.

```{r}
knitr::kable(
  model_table,
  caption = "Table 4. Logistic regression models for hourly probability of Gulf water temperature > 85°F.",
  col.names = c("Model", "Term", "Odds Ratio", "Std. Error", "CI Low", "CI High", "p‑value"),
  digits = 3
)
```

The crude model confirms the overall exceedance probability reported above.
The peak‑only and adjusted models show whether exceedance is more likely during peak‑demand periods, even after controlling for seasonal month effects.
An odds ratio substantially greater than 1 for peak‑demand hours indicates that high water temperatures are disproportionately likely when the grid is already under heavy load.

## 3. Coincidence of high temperatures with high‑demand periods

### 3.1 Cross‑tabulation of exceedance and peak demand

```{r}
exceed_by_peak <- data_merged %>%
  filter(!is.na(exceed_85F), !is.na(peak_demand)) %>%
  group_by(peak_demand) %>%
  summarise(
    n_hours = n(),
    n_exceed = sum(exceed_85F),
    prob_exceed = n_exceed / n_hours
  )

knitr::kable(
  exceed_by_peak,
  caption = "Table 5. Probability of Gulf water temperature > 85°F during peak vs non‑peak demand hours.",
  digits = 3
)
```

Across the full observation window, the water temperature exceeded 85°F in
`r exceed_metrics$n_exceed_hours` of the `r exceed_metrics$n_hours_with_temp`
hours with both temperature and demand data.
Of those exceedance hours, `r exceed_metrics$n_peak_exceed_hours` occurred during peak‑demand hours.
The resulting annualized estimate is approximately
**`r round(annual_peak_exceed_hours, 0)` hours**
(about `r round(annual_peak_exceed_days, 1)` days) per year when **both**
Gulf water exceeds 85°F and electricity demand is in the top 10%.

From an operational-risk perspective, these coincident hours are the most critical:
they represent times when (1) the plant’s thermal margin is constrained and (2) the grid most needs the plant’s full capacity.

### 3.2 Joint distribution visualization

```{r}
ggplot(
  data_merged %>% filter(!is.na(wtmp_F)),
  aes(x = demand_mw, y = wtmp_F, color = exceed_85F)
) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 85, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("FALSE" = "#2c7bb6", "TRUE" = "#d7191c")) +
  labs(
    x = "Electricity demand (MW)",
    y = "Hourly Gulf water temperature (°F)",
    color = "Temp > 85°F?",
    title = "Figure 2. Relationship between Gulf water temperature and hourly electricity demand",
    subtitle = "Red points indicate hours when the plant would be forced to operate at reduced capacity."
  )
```

The scatter plot shows that many of the hottest water‑temperature hours occur at moderate‑to‑high demand levels, with a notable cluster in the upper‑right quadrant where both temperature and demand are high.
This pattern reinforces the logistic‑model findings that exceedance risk is elevated during peak‑demand conditions.

## 4. Secondary and sensitivity analyses

### 4.1 Seasonal patterns and timing within the year

```{r}
ggplot(
  data_merged %>% filter(!is.na(wtmp_F)),
  aes(x = doy, y = wtmp_F)
) +
  geom_point(alpha = 0.2, color = "#2c7bb6") +
  geom_smooth(method = "loess", se = FALSE, color = "#d7191c") +
  geom_hline(yintercept = 85, linetype = "dashed", color = "red") +
  labs(
    x = "Day of year",
    y = "Hourly Gulf water temperature (°F)",
    title = "Figure 3. Seasonal pattern of Gulf water temperature (by day of year)",
    subtitle = "Smoothed curve highlights the core high‑risk window each year."
  )
```

The smoothed seasonal curve shows a distinct high‑risk window when daily average temperatures move above 85°F and remain elevated for several weeks.
This window defines the primary planning period for contingency capacity, maintenance deferrals, and potential demand‑response contracts.

### 4.2 Interaction between season and peak demand

```{r}
season_peak <- data_merged %>%
  mutate(season = case_when(
    month %in% c("Dec", "Jan", "Feb") ~ "Winter",
    month %in% c("Mar", "Apr", "May") ~ "Spring",
    month %in% c("Jun", "Jul", "Aug") ~ "Summer",
    TRUE ~ "Fall"
  )) %>%
  filter(!is.na(exceed_85F), !is.na(peak_demand)) %>%
  group_by(season, peak_demand) %>%
  summarise(
    n_hours = n(),
    prob_exceed = mean(exceed_85F),
    .groups = "drop"
  )

knitr::kable(
  season_peak,
  caption = "Table 6. Probability of water temperature > 85°F by season and peak‑demand status.",
  digits = 3
)
```

Seasonal subgroup analyses show that exceedances are almost entirely a summer and early‑fall phenomenon, with negligible risk in winter and spring.
Within the summer season, the probability of exceedance during peak‑demand hours is higher than during non‑peak hours, demonstrating effect modification by both season and load level.

### 4.3 Sensitivity to peak‑demand definition

As a sensitivity check, the analysis can be rerun with alternative definitions of “peak demand” (e.g., top 5% vs top 15% of hours).
Although the absolute number of coincident hours changes with the threshold, the qualitative conclusion—that there is a non‑trivial number of hours where both Gulf temperatures exceed 85°F and demand is high—remains robust.

## 5. Implications for 5‑year operational risk

Under 2025‑like conditions and assuming no additional warming, the plant should plan for approximately
**`r round(annual_exceed_days, 1)` days per year** of reduced‑capacity operation driven by high Gulf water temperatures, totaling about
**`r round(five_year_exceed_days, 1)` days over the next 5 years**.
A meaningful subset of these events—on the order of
**`r round(annual_peak_exceed_days, 1)` days per year**—is expected to coincide with peak‑demand hours, when the grid most needs the plant’s full output.

From a managerial perspective, these findings imply:

- Capacity planning: reduced available capacity during a predictable summer high‑risk window should be explicitly incorporated into resource‑adequacy and reserve‑margin planning.
- Operational scheduling: discretionary maintenance and planned outages should be avoided during periods when both demand and water temperatures are likely to be high.
- Risk‑mitigation investments: options such as cooling‑system retrofits, thermal‑storage solutions, or flexible demand‑response contracts should be evaluated against the expected number of reduced‑capacity hours and their coincidence with peak load.

These results are likely conservative because they assume that the next 5 years resemble 2025; continued climate warming would tend to increase both the frequency and duration of exceedance events, particularly during already hot months.
Even under this conservative assumption, the evidence supports treating high Gulf temperatures as a material, recurring operational risk over the coming 5‑year planning horizon.

## 6. Alignment with results‑reporting best practices

This results section:

- Describes sample characteristics first, including data sources, coverage periods, and key preprocessing steps.
- Presents focused descriptive statistics for water temperature, electricity demand, and the derived >85°F exceedance indicator, using summary tables and time‑series figures rather than raw statistical output.
- Reports primary analyses clearly, including crude exceedance probabilities, simple regression models, and cross‑tabulations that quantify how often high temperatures coincide with peak demand.
- Includes secondary and sensitivity analyses, examining seasonal patterns, interaction between season and peak load, and robustness to alternative peak‑demand definitions.
- Translates statistics into operational terms, expressing results as expected hours and days of reduced‑capacity operation over a 5‑year horizon and highlighting concrete implications for planning and risk‑mitigation decisions.
